---
title: "Beyond Prediction: A Causal Framework in R"
subtitle: "Disentangling Simpson's Paradox and Collider Bias with an In-Silico Laboratory"
author: "Kimberley (isa229)"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
  cache: false
---

```{r setup}
#| include: false
library(here)
library(dplyr)
library(broom)
library(ggplot2)
library(knitr)

# Source the framework. All numbers in this document are computed live from
# these functions, so the narrative can never drift from the code.
source(here::here("R", "01_define_dag.R"))
source(here::here("R", "02_simulate_data.R"))
source(here::here("R", "03_naive_models.R"))
source(here::here("R", "04_causal_models.R"))
source(here::here("R", "06_validate_ground_truth.R"))
source(here::here("R", "07_comparison_plots.R"))

set.seed(2026)
SIM_N <- 50000

# Generate the dataset once and reuse it throughout the document.
df <- simulate_delivery_data(n_customers = SIM_N, seed = 2026)
df_num <- df |> mutate(churn_numeric = as.numeric(as.character(churned)))

# Helper: extract the has_exception log-odds from a glm formula
logodds <- function(formula) {
  unname(coef(glm(formula, data = df_num, family = binomial()))["has_exception"])
}

# Helper: marginal ATE (percentage points) via g-computation from a glm
ate_pp <- function(formula) {
  m  <- glm(formula, data = df_num, family = binomial())
  d1 <- df_num |> mutate(has_exception = 1L)
  d0 <- df_num |> mutate(has_exception = 0L)
  mean(predict(m, d1, type = "response") - predict(m, d0, type = "response")) * 100
}

inline_pp <- function(x) sprintf("%+.1f pp", x)
```

> **Reproducibility note.** Every figure and every number in this document is
> *computed live* when the page is rendered (seed = `2026`,
> `r format(SIM_N, big.mark = ",")` simulated customers). Nothing is hard-coded.
> If you change the data-generating process, this explanation updates itself.

# The Problem: A Model That Recommends the Absurd

Imagine a logistics company that has trained a high-accuracy churn model. One of its
strongest signals is **delivery exceptions** (late or failed deliveries). The model
reports that customers who experience exceptions are *less* likely to churn.

Taken literally, the recommendation is: **slow your deliveries down to retain
customers.** No operator would ship that. So the model is an excellent *predictor* and
a terrible *decision tool* — and the gap between those two things is what causal
inference exists to close.

This document builds a small, fully-simulated "in-silico laboratory" where we **know
the true answer**, deliberately walk into two classic statistical traps, and then show
how a causal workflow recovers the truth.

# Why Prediction ≠ Causation

Predictive models answer *"given what I observe, what is likely?"* Causal questions
ask *"if I intervene and change X, what happens to Y?"* A variable can be wonderfully
predictive yet causally misleading — and conditioning on the wrong variables can flip
an estimate's sign entirely.

We engineer two such failures on purpose:

1. **Simpson's Paradox** — a confounder we must *include*.
2. **Collider bias** — a variable we must *exclude*.

# The Scenario and Its Causal Map

The business question is: **does experiencing a delivery exception cause a customer to
churn?** The data-generating process embeds the following causal structure.

```{r dag}
#| fig-cap: "The causal DAG. Order_Volume is a confounder (a backdoor path); Support_Ticket is a collider; Impatience is an unmeasured (latent) cause."
dag_results <- define_causal_dag()
dag_results$plot
```

The variables and their causal roles:

| Variable | Causal role | Observed? | Plain-English meaning |
|---|---|:---:|---|
| `has_exception` | **Treatment** | ✅ | Did the customer hit a delivery problem? |
| `churned` | **Outcome** | ✅ | Did the customer leave? |
| `order_volume` | **Confounder** | ✅ | Engagement / order frequency |
| `opened_support_ticket` | **Collider** | ✅ | Caused by exceptions *and* impatience |
| `impatience` | **Latent cause** | ❌ | A personality trait we never measure |
| `monthly_spend_usd` | Precision covariate | ✅ | Sharpens the estimate |
| age, emails, logins | Noise | ✅ | Distractors a good analyst drops |

Crucially, `impatience` is generated but **never returned in the dataset** — exactly as
in real life, where you rarely get to measure the thing you most wish you could.

## The True Effect (the thing we get to know)

Because we wrote the simulation, we know the ground truth exactly:

```{r ground-truth}
truth <- get_ground_truth_ate()
```

- The **structural effect** of an exception on churn is `r truth$log_odds` on the
  log-odds scale.
- The **true marginal Average Treatment Effect (ATE)** — the change in churn
  *probability* if everyone (vs no one) experienced an exception, computed by
  g-computation on the data-generating process — is
  **`r inline_pp(truth$ate_pp)`**.

In words: exceptions genuinely *increase* churn by roughly
`r round(truth$ate_pp, 1)` percentage points. Keep that number in mind; it's the
target every model is trying to hit.

# Trap #1: Simpson's Paradox (Confounding)

High-volume, highly-engaged customers place many orders. More orders means more
opportunities for something to go wrong, so they encounter **more** exceptions. But
those same loyal customers churn **less**. `order_volume` therefore influences both the
treatment and the outcome — the textbook definition of a confounder.

The consequence is a sign reversal between the aggregate and the strata:

```{r simpsons-fig}
#| fig-cap: "Within every order-volume decile, customers WITH a delivery exception (red) churn more than those without (blue). Yet exception customers cluster in the high-volume, low-churn deciles, so the AGGREGATE comparison reverses."
plot_simpsons_paradox(df)
```

```{r simpsons-numbers}
agg_diff <- (mean(df_num$churn_numeric[df_num$has_exception == 1]) -
             mean(df_num$churn_numeric[df_num$has_exception == 0])) * 100
mean_vol_exposed   <- mean(df$order_volume[df$has_exception == 1])
mean_vol_unexposed <- mean(df$order_volume[df$has_exception == 0])
```

The mechanism in numbers:

- **Aggregate:** exception customers appear to churn `r sprintf("%.1f", abs(agg_diff))`
  percentage points *less* — the illusion.
- **Where the exposed live:** their mean order-volume is
  `r sprintf("%+.2f", mean_vol_exposed)` versus
  `r sprintf("%+.2f", mean_vol_unexposed)` for everyone else. They sit squarely in the
  loyal, low-churn region.
- **Within each decile:** the relationship flips back to positive — the truth.

**The fix is to ADD the confounder** to the model, closing the backdoor path.

# Trap #2: Collider Bias

A support ticket is opened when a customer is annoyed enough to complain. Two things
drive that: a delivery **exception**, and the customer's underlying **impatience**.
Because the ticket has two arrows pointing *into* it, it is a **collider**.

Conditioning on a collider opens a non-causal path between its parents, manufacturing a
spurious association between `has_exception` and (via impatience) `churn`. And here's
the cruel twist: the ticket is also the single most *predictive* feature of churn —
precisely why a predictive-first workflow will keep it, and precisely why doing so
poisons the causal estimate.

**The fix is to REMOVE the collider** — never adjust for it.

# The In-Silico Laboratory

The reason we can *prove* any of this is that the data is simulated. We know the true
ATE, so we can grade each modeling choice against it. Let `dagitty` tell us, purely
from the assumptions encoded in the DAG, which variables we must adjust for:

```{r adjustment-set}
adj <- get_adjustment_strategy(dag_results$dag)
adj
```

```{r adjustment-vars}
#| include: false
adj_vars <- unlist(adj)
```

The minimal sufficient adjustment set is **{ `r paste(adj_vars, collapse = ", ")` }** —
it includes the confounder and, importantly, it does **not** include the collider.

# The Two Recipes

Both pipelines use the same `tidymodels` machinery; they differ only in which variables
they keep.

**The naive (predictive-first) recipe** drops the confounder (`order_volume` looked
"irrelevant") and keeps the collider (`opened_support_ticket` was too predictive to
throw away):

```{r naive-recipe}
#| eval: false
recipe(churned ~ ., data = train) |>
  update_role(customer_id, new_role = "ID") |>
  step_rm(order_volume) |>          # drops the confounder  (mistake #1)
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
# opened_support_ticket is RETAINED  (mistake #2)
```

**The DAG-guided causal recipe** does the opposite — adds the confounder, drops the
collider and the noise:

```{r causal-recipe}
#| eval: false
recipe(churned ~ ., data = train) |>
  update_role(customer_id, new_role = "ID") |>
  step_rm(opened_support_ticket,    # drops the collider   (the key move)
          customer_age_years, marketing_emails_clicked, app_logins_last_7_days) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
# order_volume is RETAINED
```

# The Reveal

```{r estimates}
true_ate   <- truth$ate_pp
naive_ate  <- ate_pp(churn_numeric ~ has_exception + opened_support_ticket + monthly_spend_usd)
causal_ate <- ate_pp(churn_numeric ~ has_exception + order_volume + monthly_spend_usd)

results <- tibble(
  Method = c("Ground Truth (DGP)",
             "Naive model (omit confounder, keep collider)",
             "Causal model (add confounder, drop collider)"),
  `ATE (pp)` = c(true_ate, naive_ate, causal_ate),
  `Bias (pp)` = c(0, naive_ate - true_ate, causal_ate - true_ate),
  Direction = c("Increases churn",
                ifelse(naive_ate > 0, "Increases churn", "DECREASES churn (WRONG)"),
                ifelse(causal_ate > 0, "Increases churn", "Decreases churn"))
)
kable(results, digits = 1, caption = "Marginal ATE: the naive model gets the sign wrong; the causal model recovers the truth.")
```

```{r money-shot}
#| fig-cap: "The money shot. The naive estimate (red) sits on the wrong side of zero; the causal estimate (blue) lands on the true effect (green line)."
comparison_tbl <- tibble(
  Method = c("Ground Truth (DGP)", "Naive Model (Simpson + Collider)", "Causal Model (DAG-Guided)"),
  ATE_pp = c(true_ate, naive_ate, causal_ate)
)
plot_ground_truth_comparison(comparison_tbl)
```

The naive model concludes exceptions **`r ifelse(naive_ate < 0, "reduce", "increase")`**
churn (`r inline_pp(naive_ate)`) — the absurd recommendation we started with. The
causal model recovers **`r inline_pp(causal_ate)`**, essentially the true
`r inline_pp(true_ate)`.

## How the Estimate Moves Across Specifications

It is instructive to watch the `has_exception` coefficient migrate as we change the
adjustment set:

```{r trajectory}
#| fig-cap: "Only the DAG-guided specification (add confounder, drop collider) lands near the true structural coefficient."
plot_coefficient_trajectory(df)
```

```{r trajectory-numbers}
traj <- tibble(
  Specification = c("Aggregate (exception only)",
                    "Naive ML (+ collider + spend)",
                    "DAG-guided (+ confounder + spend)",
                    "Over-adjusted (DAG + collider)"),
  `log-odds` = c(
    logodds(churn_numeric ~ has_exception),
    logodds(churn_numeric ~ has_exception + opened_support_ticket + monthly_spend_usd),
    logodds(churn_numeric ~ has_exception + order_volume + monthly_spend_usd),
    logodds(churn_numeric ~ has_exception + order_volume + monthly_spend_usd + opened_support_ticket)
  )
)
kable(traj, digits = 3, caption = "has_exception log-odds across specifications.")
```

# An Honest Accounting

A causal demo is only worth as much as its honesty about its own assumptions.

## Why we headline the ATE, not the logistic coefficient

```{r noncollapsibility}
causal_logodds <- logodds(churn_numeric ~ has_exception + order_volume + monthly_spend_usd)
```

The DAG-guided **log-odds** coefficient is `r sprintf("%.3f", causal_logodds)`, a little
below the structural `r truth$log_odds`. This is **not bias** — it is the well-known
**non-collapsibility** of the odds ratio: when a strong cause of the outcome
(`impatience`) is unmeasured, the conditional odds ratio shrinks toward the null even
when the model is correctly specified.

The **marginal ATE on the probability scale is collapsible**, so it recovers the truth
(`r inline_pp(causal_ate)` vs `r inline_pp(true_ate)`). It is also the number a business
actually wants — *"if we eliminated exceptions, churn would fall by about this much."*
That's why it's our headline estimand.

## Uncertainty: a bootstrap confidence interval

```{r bootstrap}
boot <- bootstrap_ate_ci(df, n_bootstrap = 500, seed = 2026)
```

A percentile bootstrap gives a 95% confidence interval for the causal ATE of
**[`r sprintf("%.1f", boot$CI_Lower)`, `r sprintf("%.1f", boot$CI_Upper)`] pp** — comfortably
away from zero and bracketing the true effect.

## Robustness: how strong an unmeasured confounder would have to be

Since `impatience` is exactly the kind of variable we can never measure, we quantify the
finding's fragility with an **E-value** (VanderWeele & Ding, 2017):

```{r evalue}
or_causal <- exp(causal_logodds)
ev <- compute_evalue(or_causal)
```

With a causal odds ratio of `r sprintf("%.2f", or_causal)`, the E-value is
**`r sprintf("%.2f", ev$evalue)`**. In plain terms: an unmeasured confounder would need
to be associated with *both* the exception and churn by a factor of at least
`r sprintf("%.2f", ev$evalue)`× — beyond everything already adjusted for — to explain the
effect away. That is a demanding bar, which gives us confidence in the result.

# Reproduce This Yourself

```r
# 1. Activate the environment (exact package versions via rv)
source("rv/scripts/activate.R")

# 2. Regenerate every figure and the validation output
source("R/05_generate_figures.R")

# 3. Run the test suite (asserts both traps emerge and the causal model wins)
testthat::test_dir("tests/testthat")

# 4. Re-render this document
quarto::quarto_render("docs/causal-framework-explained.qmd")
```

Everything is seeded (`seed = 2026`), so your results will match these exactly.

# Takeaways

1. **A great predictor can give terrible advice.** Predictive importance is not causal truth.
2. **Draw the DAG first.** Encoding assumptions makes the right adjustment set a mechanical question — answered here by `dagitty`.
3. **Confounders go IN, colliders stay OUT.** The two mistakes pushed the estimate the wrong way; the fixes are opposite in direction.
4. **Simulate to validate.** An in-silico laboratory is the only place you can grade a causal method against a known truth.
5. **Report decision-relevant, honest effects.** Prefer the collapsible marginal ATE, attach a CI, and stress-test with an E-value.

# References

- Pearl, J. (2009). *Causality: Models, Reasoning, and Inference.*
- Hernán, M.A. & Robins, J.M. (2020). *Causal Inference: What If.*
- VanderWeele, T.J. & Ding, P. (2017). "Sensitivity Analysis in Observational Research: Introducing the E-Value." *Annals of Internal Medicine.*
- Elwert, F. & Winship, C. (2014). "Endogenous Selection Bias: The Problem of Conditioning on a Collider Variable."
- Greenland, S., Robins, J.M., & Pearl, J. (1999). "Confounding and Collapsibility in Causal Inference."
